Synthesis of Linear Functional Programs

### 2021-05-07 article/article.tex
tweaks

\mypara{Goals.}
 % Goals
In the end, we intend to be able to do full and partial synthesis of
well-typed programs. Full synthesis consists of the production of a
function (or set of) satisfying the specification; Partial synthesis
is the ``completion'' of a partial program (i.e. a function with a
\emph{hole} in it). The work will start from a small core linear --
i.e. \emph{resource-aware} functional language, building up to
recursive types/functions, with potential other avenues of further
extension. 
%
We will evaluate the work through
expressiveness benchmarks (i.e. can we synthesize ``X''?) and time
measurements (i.e. how fast can we synthesize ``X''?').

@@

\todo{Merge! These 2 paragraphs}
A type system can be formally described through a set of inference
rules that inductively define a judgment of the form $\Gamma \vdash M
: A$, stating that program expression $M$ has type $A$ according to
the \emph{typing assumptions} for variables tracked in $\Gamma$.
\dots

@@

A type system can be formally described through natural deduction -- as a set of rules that determines the concept of deduction for some language~\cite{prawitznd65}.
This comes from the close relation between logic and programming languages, the Curry-Howard isomorphism or correspondence. A logical proof describes a program, and a program describes a proof.
The main observations made are ``propositions as types'', ``proofs as
programs'', ``proof checking as type checking''
\cite{https://www.cs.cmu.edu/~fp/courses/15312-f04/handouts/23-curryhoward.pdf}.
Formalizing our type system, the language i.e. the propositions of the logic are the types themselves, the inference rules are used to construct a proof which corresponds directly to a program, and the correctness of the proof assures the program is well-typed.
The typical natural deduction judgment
\[
    A \textrm{ is a proposition}
\]
is extended to include a \emph{term}
(i.e. to make proofs explicit with a formal definition -- these proofs map directly to a $\lambda$-calculus program as stated by the Curry-Howard correspondence????),
and modified to include a list of hypothesis. 
This extended judgment, called?? a typing judgment, has the form:
\[
    \Gamma \vdash M : A
\]
In the type system context, the list of hypothesis, $\Gamma$, is called \emph{typing environment} (i.e. set of variables and their types),
 the term $M$ is the \emph{program}, and $A$, the natural deduction proposition, is the \emph{type}.
 This judgment can be read -- $M$ has type $A$ in $\Gamma$.

Type rules assert the validity of certain judgments on the basis of other judgments already known to be valid \cite{http://homepage.cs.uiowa.edu/~tinelli/classes/185/Fall06/notes/cardelli-95.pdf}.
We introduce a few typing i.e. inference rules to display how a type system would be formalized.
In natural deduction, there are both \emph{introduction} and \emph{elimination} rules for almost all logical connectives.
The following rule is a hypothesis rule ??, não sei muito bem como sair daqui:)
\[
    \infer*[right=(var)]
    {  }
    {\Gamma, u: A \vdash u : A}
\]
It reads -- the program $u$ has type $A$ if our typing environment holds a variable labeled $u$ of type $A$.
[também não sei muito bem onde quero chegar com isto :)]
As another concrete example, let us describe the inference rule for the function type ($A \supset B$):
\[
    \infer*[right=($\supset I$)]
    {\Gamma, u: A \vdash M : B}
    {\Gamma \vdash \lambda u. M : A \supset B }
\]
It reads -- the program $\lambda u. M$ has type $A \supset B$ ($A$ implies $B$) as long as $M$ is a program of type $B$ knowning that $u$ is a variable of type $A$.
The rule shown above is an \emph{introduction} rule, that introduces the ``implication'' connective,

% Linear Logic
\subsection{Linear Logic}

@@

@@

More about the concepts than citing concrete works (but citations are
needed!):
\begin{enumerate}
\item PL/Type systems in a ``formal'' sense (inference rules, etc)
\item Propositions as types: Proofs as programs, proof search as
  synthesis, etc.
\item Linear logic: a ``resource aware logic''
\item Proof search in LL: challenges, focusing as a ``solution''
\item Linear $\lambda$-calculus: what does it look like.
\item Type-based synthesis: ``inverting'' a type system.
\end{enumerate}

@@

Concrete works:
\begin{itemize}
\item Synthesis for graded types (no recursion   -- in that work, no
  obvious connection of grading with
  concurrency)~\cite{DBLP:conf/lopstr/HughesO20}

@@

@@

\item Synthesis of heap-manipulating
  programs~\cite{DBLP:journals/pacmpl/PolikarpovaS19}

Less related, but more evidence of this idea of turning ``checking''
systems into synthesis frameworks.

\item Resource-guided synthesis \cite{DBLP:conf/pldi/KnothWP019}

This resource-guided means something a bit different. Programs satisfy
a functional specification and a symbolic resource bound in the sense
of amortized analysis, but can provide technical insights.

>>>>

\mypara{Goals}
 % Goals
In the end, we intend to be able to do full and partial synthesis of
well-typed programs. Full synthesis consists of the production of a
function (or set of) satisfying the specification; Partial synthesis
is the ``completion'' of a partial program (i.e. a function with a
\emph{hole} in it). The work will start from a small core linear --
i.e. \emph{resource-aware} functional language, building up to
recursive types/functions, with potential other avenues of further
extension. 
%
We will evaluate the work through
expressiveness benchmarks (i.e. can we synthesize ``X''?) and time
measurements (i.e. how fast can we synthesize ``X''?').

@@

\mypara{Type Systems} A type system can be formally described through
a set of inference rules that inductively define a judgment of the
form $\Gamma \vdash M : A$, stating that program expression $M$ has
type $A$ according to the \emph{typing assumptions} for variables
tracked in $\Gamma$. For instance,
$x{:}\mathsf{Int}, y{:}\mathsf{Int} \vdash x+y : \mathsf{Int}$ states
that $x+y$ has type $\mathsf{Int}$ under the assumption that $x$ and
$y$ have type $\mathsf{Int}$.  An expression $M$ is deemed well-typed
with a given type $A$ if one can construct a typing derivation with $M :
A$ as its conclusion, by repeated application of the inference rules.

The simply-typed $\lambda$-calculus is a typed core functional
language~\cite{} that captures \dots. Its syntax consists of
functional abstraction, written $\lambda x{:}A.M$, denoting the
an (anonymous) function that takes an argument of type $A$, bound to
$x$ in $M$; and application $M\,N$, with the standard meaning, and
variables $x$. For instance, the \dots \todo{Identity function!}

\mypara{Propositions as Types}
%
It turns out that the inference rules of the simply-typed
$\lambda$-calculus are closely related to those of a system of natural
deduction for intuitionistic logic~\cite{prawitznd65}. This
relationship, known as the Curry-Howard correspondence~\cite{},
identifies that the propositions of intuitionistic logic can be read
as types for the simply-typed $\lambda$-calculus, and their proofs are
exactly the \dots.
The main observations made are ``propositions as types'', ``proofs as
programs'', ``proof checking as type checking''~\cite{}.

Inference in ND are categorized as introduction or elimination, these
correspond to constructors and destructors in programming languages
(e.g. function abstraction vs application, construction of a pair vs
projection).

We introduce a few typing rules to display how a type system can be
formalized.  In natural deduction, there are both \emph{introduction}
and \emph{elimination} rules for almost all logical
connectives.\todo{se mencionares intro e elim antes nao precisas de
  repetir aqui.}  The following rule captures the nature of a
hypothetical judgment, allowing for reasoning from assumptions:
\[
    \infer*[right=(var)]
    {  }
    {\Gamma, u: A \vdash u : A}
  \]
  When seen as a typing rule for the $\lambda$-calculus, it
  corresponds to the rule for typing variables -- variable $u$ has
  type $A$ if the typing environment contains a variable $u$ of
  type $A$.

  As another concrete example, let us consider the rule for
  implication ($A \supset B$):
\[
    \infer*[right=($\supset I$)]
    {\Gamma, u: A \vdash M : B}
    {\Gamma \vdash \lambda u. M : A \supset B }
  \]
  Logically, the rule states that to prove $A\supset B$, we assume $A$
  and prove $B$. Through the Curry-Howard correspondence, implication
  corresponds to the function type -- the program $\lambda u. M$ has
  type $A \supset B$, provided $M$ has type
  type $B$ under the assumption that $u$ is a variable of type $A$.  The rule
  shown above is an \emph{introduction} rule, that introduces the
  ``implication'' connective. 
%$\rightarrow$ $\Rightarrow$, $\Longrightarrow$, $\longrightarrow$

  \todo[inline]{Mesma historia para a aplicacao/modus ponens.
Um bocadinho das ``consequencias'' do Curry-Howard. }

The Curry-Howard correspondence generalizes beyond the simply-typed
$\lambda$-calculus and propositional intuitionistic logic.
\todo{Mencionar polimorfismo corresponde a logica de ordem superior,
  etc. E tambem para logica linear \dots}

@@

% Linear Logic
\mypara{Linear Logic}

@@

\todo{Through the Curry-Howard correspondence, we can view the process
  of finding a proof of a proposition as the process of synthesizing a
program of a given type.}

@@

More about the concepts than citing concrete works (but citations are
needed!):
\begin{enumerate}
\item PL/Type systems in a ``formal'' sense (inference rules, etc) [DONE]
\item Propositions as types: Proofs as programs, proof search as
  synthesis, etc. [MOSTLY DONE]
\item Linear logic: a ``resource aware logic'' [ONGOING]
\item Proof search in LL: challenges, focusing as a ``solution'' [ONGOING]
\item Linear $\lambda$-calculus: what does it look like.
\item Type-based synthesis: ``inverting'' a type system.
\end{enumerate}

@@

\todo[inline]{Os~\cite{DBLP:conf/pldi/OseraZ15,DBLP:conf/popl/FrankleOWZ16}
  como ``estrategia geral'' de ``ler'' um sistema de tipos +
  engenharia como um sistema de sintese de programas tipificados;
  Aqui~\cite{DBLP:conf/pldi/PolikarpovaKS16} estamos num contexto
  ``mais avançado'' (polimorfismo e tipos refinados -- ``logica de
  primeira ordem'') -- tipos ricos -- trabalho aspiracional,
  tratamento de funcao recursivas, tratamento de tipos de dados
  indutivos/algebricos; Aqui~\cite{DBLP:conf/lopstr/HughesO20},
  sintese para graded modal types, tem a ver com ``linearidade'' kind
  of, abordagem baseado em focusing (portanto Curry-Howard++), no caso
  deles acaba por ser uma lógica ``modal'' vs linear -- não permite
  futuramente explorar relações com e.g. concorrência, análise de
  recursos.}

Concrete works
\begin{itemize}

@@

  \item Synthesis for graded types (no recursion   -- in that work, no
  obvious connection of grading with
  concurrency)~\cite{DBLP:conf/lopstr/HughesO20}

@@

% \item Synthesis of heap-manipulating
%   programs~\cite{DBLP:journals/pacmpl/PolikarpovaS19}

% Less related, but more evidence of this idea of turning ``checking''
% systems into synthesis frameworks.

% \item Resource-guided synthesis \cite{DBLP:conf/pldi/KnothWP019}

% This resource-guided means something a bit different. Programs satisfy
% a functional specification and a symbolic resource bound in the sense
% of amortized analysis, but can provide technical insights.
